---
title: "blog"
output: html_document
date: "2026-03-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Goal

Simulate a cointegrated relationship between air temperature (driver) and ground temperature (response) across five sensors. Ground logs have missing values that vary by sensor to mimic faulty data loggers. The simulation follows a VECM-style error-correction process so that ground temperature tracks a long-run equilibrium with air, while short-run changes in air also affect ground.

## How The Simulation Works

- Air process (`A_t`): integrated AR(1) with higher variance plus mild seasonality; air varies more than ground.
- Long-run equilibrium: `G_t` tends to `c_i + beta * A_t` with `beta < 1` so the ground amplitude is lower than air.
- Error correction: deviation `z_{t-1} = G_{t-1} - beta * A_{t-1} - c_i` is pulled back by `alpha_i < 0` (slower correction).
- Lagged short-run: ground responds to past air changes, `ΔA_{t-L}` (we use L = 7 days), making ground move sluggishly.
- Heterogeneity: sensor-level random effects for `alpha_i`, `gamma_i`, and `c_i`.
- Missingness: only ground logs are missing; each sensor gets its own missing rate, with block gaps added for the worst loggers.

The data and plot are created in `script.R`.

```{r simulate, message=FALSE, warning=FALSE}
source("script.R")
```

### Missingness By Sensor

```{r miss-table, echo=FALSE}
library(dplyr)
library(knitr)
sim %>%
  group_by(sensor) %>%
  summarize(miss_rate = mean(is.na(ground_obs))) %>%
  knitr::kable(digits = 2, caption = "Share of missing ground observations by sensor")
```

## Visual Check

The plot below shows complete air (blue) and observed ground (red) per sensor. Ground has lower amplitude, higher inertia, and lags behind air; better sensors have fewer gaps.

```{r plot, echo=FALSE, fig.width=8, fig.height=10, message=FALSE, warning=FALSE}
p
```

## Stan Model (Imputation)

We fit a hierarchical error-correction (VECM-style) state-space model in Stan to impute missing ground temperatures and quantify uncertainty. Air temperature is treated as a known covariate.

### Model Specification

Let sensors `i = 1, ..., S` and times `t = 1, ..., T`. Air temperature is `A_t` and its change is `\Delta A_t = A_t - A_{t-1}`.

- Cointegrating equilibrium per sensor:

$$G^*_{t,i} = c_i + (\beta_0 + \beta_i)\,A_t$$

- Error-correction term:

$$z_{t-1,i} = G_{t-1,i} - (\beta_0 + \beta_i)\,A_{t-1} - c_i$$

- State evolution (short-run dynamics are contemporaneous in the Stan model):

$$G_{t,i} = G_{t-1,i} + \alpha_i\,z_{t-1,i} + \gamma_i\,\Delta A_t + \varepsilon_{t,i}$$

equivalently

$$\Delta G_{t,i} = \alpha_i\,z_{t-1,i} + \gamma_i\,\Delta A_t + \varepsilon_{t,i},\quad \varepsilon_{t,i} \sim \mathcal{N}(0,\sigma_{\text{eps}})$$

- Initial state prior (sub-zero starts):

$$G_{1,i} \sim \mathcal{N}(\mu^{(G1)}_i,\,\sigma_{\text{init}}),\quad \mu^{(G1)}_i \sim \mathcal{N}(-5,\,3)$$

- Hierarchical priors for sensor effects and hyperparameters:

$$\alpha_i \sim \mathcal{N}(\mu_\alpha,\,\sigma_\alpha),\qquad \gamma_i \sim \mathcal{N}(\mu_\gamma,\,\sigma_\gamma),\qquad c_i \sim \mathcal{N}(\mu_c,\,\sigma_c)$$

$$\mu_\alpha \sim \mathcal{N}(-0.3,0.3),\quad \mu_\gamma \sim \mathcal{N}(0.6,0.3),\quad \mu_c \sim \mathcal{N}(0,2)$$

Partial pooling for cointegration slope:

$$\beta_0 \sim \mathcal{N}(0.8,0.2),\qquad \beta_i \sim \mathcal{N}(0,\,\sigma_\beta),\qquad \sigma_\beta \sim \text{Exponential}(1)$$

$$\sigma_\alpha,\sigma_\gamma,\sigma_c,\sigma_{\text{init}},\sigma_{\text{eps}},\sigma_{\text{obs}} \sim \text{Exponential}(1)$$

- Observation model (known indexes `(t_n, i_n)` for observed entries `n = 1, ..., N_{\text{obs}}`):

$$y^{\text{obs}}_n \sim \mathcal{N}(\,G_{t_n,i_n},\,\sigma_{\text{obs}}\,)$$

- Missing data (Stan slicing; unknowns `y^{\text{mis}}_m` for missing indexes `(\tau_m, j_m)`, `m = 1, ..., N_{\text{mis}}`):

$$y^{\text{mis}}_m \sim \mathcal{N}(\,G_{\tau_m,j_m},\,\sigma_{\text{obs}}\,)$$

This follows Stan’s recommended pattern: observed and missing outcomes are split and modeled separately, with missing outcomes declared as parameters. Air `A_t` and `\Delta A_t` are treated as known covariates.

Note: Our simulation uses a 7-day lag for short-run effects (`\Delta A_{t-7}`) to make ground respond more slowly. The current Stan fit uses contemporaneous `\Delta A_t`. If desired, we can replace `\Delta A_t` with `\Delta A_{t-L}` to mirror the simulation.

```{r stan-fit, message=FALSE, warning=FALSE, results='hide'}
library(cmdstanr)
library(dplyr)
library(tidyr)
library(stringr)

# prepare indices for observations
sensors <- unique(sim$sensor)
sensor_index <- setNames(seq_along(sensors), sensors)

T_ <- max(sim$t)
S_ <- length(sensors)

air_ts <- sim %>%
  filter(sensor == sensors[1]) %>%
  arrange(t) %>%
  pull(air)

dA_ts <- sim %>%
  filter(sensor == sensors[1]) %>%
  arrange(t) %>%
  pull(dA)

obs_df <- sim %>%
  filter(!is.na(ground_obs)) %>%
  mutate(i = sensor_index[sensor]) %>%
  select(t, i, y = ground_obs) %>%
  arrange(i, t)

mis_df <- sim %>%
  filter(is.na(ground_obs)) %>%
  mutate(i = sensor_index[sensor]) %>%
  select(t, i) %>%
  arrange(i, t)

stan_data <- list(
  T = T_, S = S_, A = air_ts, dA = dA_ts,
  N_obs = nrow(obs_df),
  obs_t = obs_df$t,
  obs_i = obs_df$i,
  y_obs = obs_df$y,
  N_mis = nrow(mis_df),
  mis_t = if (nrow(mis_df) > 0) mis_df$t else integer(0),
  mis_i = if (nrow(mis_df) > 0) mis_df$i else integer(0)
)

mod <- cmdstan_model("vecm_partial_pooling.stan")
fit <- mod$sample(
  data = stan_data,
  seed = 100,
  
  chains = 1, parallel_chains = 4,
  iter_warmup = 800, iter_sampling = 800,
  adapt_delta = 0.9
)

# summarize latent ground G[t,i]
sum_G <- fit$summary(variables = c("G"))
sum_G <- sum_G %>%
  select(variable, mean, sd) %>%
  mutate(idx = str_match(variable, "G\\[(\\d+),(\\d+)\\]")) %>%
  mutate(t = as.integer(idx[,2]), i = as.integer(idx[,3])) %>%
  select(t, i, ground_mean = mean, ground_sd = sd)

imputed <- tidyr::expand_grid(t = 1:T_, i = 1:S_) %>%
  left_join(sum_G, by = c("t","i")) %>%
  mutate(sensor = names(sensor_index)[i]) %>%
  select(t, sensor, ground_mean, ground_sd)
```

### Plot With Imputations

The plot overlays the posterior mean ground temperature and highlights uncertainty (±2 SD) at time points where ground data were missing.

```{r imputed-plot, echo=FALSE, fig.width=8, fig.height=11, message=FALSE, warning=FALSE}
library(ggplot2)

plot_df <- sim %>%
  left_join(imputed, by = c("t","sensor")) %>%
  mutate(is_missing = is.na(ground_obs))

gg <- ggplot() +
  geom_line(data = plot_df, aes(t, air, color = "Air"), linewidth = 0.5) +
  geom_line(data = plot_df, aes(t, ground_obs, color = "Ground (obs)"), linewidth = 0.5, na.rm = TRUE) +
  geom_line(data = plot_df, aes(t, ground_mean, color = "Ground (mean)"), linewidth = 0.6, alpha = 0.9) +
  geom_linerange(
    data = dplyr::filter(plot_df, is_missing),
    aes(x = t, ymin = ground_mean - 2 * ground_sd, ymax = ground_mean + 2 * ground_sd, color = "Imputed ±2 SD"),
    alpha = 0.7,
    linewidth = 0.7
  ) +
  facet_wrap(~ sensor, ncol = 1, scales = "free_y") +
  scale_color_manual(values = c("Air" = "#1f77b4", "Ground (obs)" = "#d62728", "Ground (mean)" = "#8c1515", "Imputed ±2 SD" = "#d62728"), name = NULL) +
  labs(title = "Air vs Ground (Observed and Imputed)", x = "Time", y = "Temperature") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")

gg
```

```{r save-plots, echo=FALSE, message=FALSE, warning=FALSE}
# Save both the simulated and imputed plots to the project public path
output_dir <- file.path("..", "..", "..", "public", "blogs", "2026-03-04-spring-thaw")
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

ggsave(filename = file.path(output_dir, "simulated-ground-air.png"), plot = p, width = 8, height = 10, dpi = 120)
ggsave(filename = file.path(output_dir, "imputed-ground-air.png"), plot = gg, width = 8, height = 11, dpi = 120)
```

## Notes

- Sensors start below 0°C in the simulation; the Stan model also places a prior for sub-zero initial states.
- Missing data are handled per Stan's recommendations by separating observed and missing indices and introducing `y_mis` as parameters linked to the latent state.
- The tibble `sim` has: `t`, `sensor`, `air`, `dA`, `ground_true`, `ground_obs`. The `imputed` tibble adds `ground_mean`, `ground_sd` from the posterior.
