---
title: "Getting Quarto Started in NextJS"
date: 01/02/2024
date-format: "MMM D, YYYY"
format:
  html:
    embed-resources: true
editor: visual
---

## Overview

This guide demonstrates an initial approach to displaying Quarto documents in a NextJS application using iframes. I am writing this post because I couldn't find any solution to this issue on the interwebs.

While not the final solution, it provides a working foundation for integrating Quarto's notebooks with NextJS's modern web framework capabilities. The end goal is to move beyond iframes to a fully integrated solution where Quarto content is parsed into native React components, but this implementation serves as a practical starting point for those looking for a quick and dirty solution

## File Structure

    app/
    ├── blog/
    │   ├── [slug]/
    │   │   ├── layout.tsx  # Right sidebar with blog links
    │   │   └── page.tsx    # iframe container for Quarto content
    │   ├── BlogSummary.tsx # Blog metadata and post info
    │   └── page.tsx        # Main blog listing page
    └── public/
        └── blogs/          # Quarto HTML files

## Implementation Steps

### 1. Quarto Document Setup

Create your Quarto document with embedded resources enabled:

``` yaml
---
title: "Your Blog Title"
embed-resources: true
---
```

### 2. Generate Self-Contained HTML

Run Quarto's render command to generate a self-contained HTML file:

``` bash
quarto render document.qmd
```

### 3. NextJS Integration

Place the generated HTML in your public folder:

``` bash
mv document.html public/blogs/
```

### 4. Create Blog Components

#### Layout Component (blog/\[slug\]/layout.tsx)

``` typescript
export default function BlogLayout({
  children,
}: {
  children: React.ReactNode
}) {
  return (
    <div className="flex h-screen m-0 p-0">
      <main className="flex-1 m-0 p-0">
        {children}
      </main>
      <aside className="w-48 bg-white shadow-lg h-screen overflow-y-auto border-l">
        <div className="p-3">
          <h2 className="text-lg font-semibold mb-3">All Posts</h2>
          <nav>
            {BlogPosts.map((post) => (
              <a key={post.title} href={post.link}>
                {post.title}
              </a>
            ))}
          </nav>
        </div>
      </aside>
    </div>
  );
}
```

#### Page Component (blog/\[slug\]/page.tsx)

``` typescript
interface PageProps {
    params: Promise<{
      slug: string;
    }>;
  }
  
  export default async function BlogPost({ params }: PageProps) {
    const { slug } = await params;
    
    return (
      <div className="w-full h-screen">
        <iframe
          src={`/blogs/${slug}.html`}
          className="w-full h-full border-none"
          title="Blog post content"
          style={{
            minHeight: '100vh',
            width: '100%',
            margin: 0,
            padding: 0,
            border: 'none'
          }}
        />
      </div>
    );
  }
```

### 5. Blog Metadata

Update BlogSummary.tsx with your post information:

``` typescript
interface BlogPost {
  title: string;
  description: string;
  link: string;
  type: 'blog' | 'project';
  tags: string[];
  featured: string;
  image: string;
  linkType?: 'email' | 'github' | 'external';
}

export const BlogPosts: BlogPost[] = [
  {
    title: "Getting Quarto Started in NextJS",
    description: "How to integrate Quarto with NextJS",
    link: "/blog/quarto-nextjs",
    type: "blog",
    tags: ["nextjs", "web dev", "quarto"],
    featured: 'true',
    image: "/favicon.ico"
  }
];
```

## Benefits of This Approach

-   Maintains Quarto's original styling and functionality
-   Simple implementation with minimal configuration
-   Clean separation between Quarto content and NextJS app
-   Easy to update and manage blog posts
-   Avoids having a .html URL tag

## Future Steps

-   Implement a custom React component parser to directly process .qmd files
-   Extract and transform Quarto HTML elements into optimized React components
-   Integrate Quarto's JavaScript runtime with NextJS's client-side rendering
-   Leverage NextJS's built-in Image and Script optimization for better performance
-   Utilize Server Components for improved SEO and reduced client-side JavaScript
-   Implement static generation for Quarto content during build time
-   Create a custom webpack loader for direct .qmd file importing

## Example Quarto Output

## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).

```{r}
library(ggplot2)

x = seq(1:100)
y = seq(1:100)^2
df = data.frame(x,y)

ggplot(df,aes(x = x, y = y))+geom_point()

```

```{r}
library(ggplot2)
library(plotly)

x = seq(1:100)
y = seq(1:100)^2
df = data.frame(x,y)

p = ggplot(df,aes(x = x, y = y))+geom_point()

p = ggplotly(p)

p
```

this is my $s$ and this is my $\phi$
